---
title: "**A Case Study on the Impact of Microfinance Institutions on Poverty Alleviation
  in Low-Income Households in the Philippines, 2021-2024**"
author: "U. Cortes, W.E. Limguangco"
date: "2025-06-10"
output:
  html_document: default
  pdf_document: default
css: style.css
bibliography: references.bib
csl: apa.csl
---

## *Abstract* 

## *1 Background & Context* 
Poverty is one of the main issues the Philippines is experiencing to the present day. In 2021, the poverty rate was 18.1 percent, equating to  19.99 million Filipinos, and the poverty incidence among Filipino families was 13.2 percent [@PSA2022Poverty]. Furthermore, the PSA reported that the proportion of Filipino families with insufficient income to meet basic food needs stood at 3.9 percent, and among Filipinos, 6 in 100 lived below the food threshold in 2021. In 2022, the PSA has not released the official poverty statistics of the Philippines for the year, however, according to [@Domingo2022] of Inquirer, the poverty incidence in the Philippines was projected to ease to 17.1 percent from 18.1 percent in 2021, thanks to recovering labor market indicators according to the World Bank.

In 2023, the poverty rate in the Philippines dropped to 15.5 percent, equating to 17.54 million Filipinos, and the poverty incidence among families at 10.9 percent, translating to 2.99 million Filipino families whose incomes are below the poverty threshold [@PSA2024Poverty]. According to an article from [@pna_article_1235008], poverty in the Philippines is expected to decline, but with risks of extreme climatic events. Furthermore, the article stated that the poverty incidence rate for 2024 is projected to decrease to 13.6 percent. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)

poverty_data <- data.frame(
  Year = c("2021", "2022 (Estimation)", "2023", "2024 (Estimation)"),
  Poverty_Incidence = c("18.1", "17.1", "15.5", "13.6")
)

kable(poverty_data, col.names = c("Years", "Poverty Incidence (%)"), align = "c", caption = "Table 1. Poverty Incidences in the Philippines 2021 to 2024.") %>%
  kable_styling(full_width = FALSE, position = "center") %>%
  column_spec(1, width = "6cm") %>%  # Narrowing the first column
  column_spec(2, width = "6cm") %>%  # Narrowing the second column
  add_footnote("*Sources: Philippine Statistics Authority (2022 & 2024), Domingo (2022) via Inquirer, Philippine News Agency (2024).*", notation = "none")

```
The microfinance sector has experienced significant growth in the Philippines in recent years.  As of March 2022, the Bangko Sentral ng Pilipinas (BSP) reported that the microfinance sector has expanded to a loan portfolio of Php 406 billion . There are approximately 6,183 microfinance institutions (MFIs) in the Philippines. The role of MFIs in inclusion and poverty reduction plays an important role in the Philippines. [@cafamerica_microfinance_philippines_2024] further discussed the role of MFIs, such as providing financial services to low-income individuals and small businesses and promoting financial inclusion by ensuring marginalized communities have access to financial resources. Quick and Toledo explained further that providing financial resources and skills empowers individuals, especially women and vulnerable groups, strengthening social empowerment as a whole. Furthermore, the authors also explain that microfinance helps in resilience building, particularly in disaster-prone regions, as MFIs assist communities in rebuilding by providing them with essential financial tools. 


The microfinance sector has experienced significant growth in the Philippines in recent years.  As of March 2022, the Bangko Sentral ng Pilipinas (BSP) reported that the microfinance sector has expanded to a loan portfolio of Php 406 billion [@batayola_martir_2024]. There are approximately 6,183 microfinance institutions (MFIs) in the Philippines [@rapisura2023philippine]. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
library(scales)  # For number formatting

microfinance_data <- data.frame(
  Microfinance_Institutions = c("Banks (as of 2022)", "Cooperatives (2020)", "Microfinance NGOs (as of 2022)", 
                                "Financing Companies (as of 2017)", "Lending Companies (as of 2017)", "Total"),
  Number_of_Institutions = c(139, 2762, 27, 673, 2582, 6183)
)

# Format numbers with commas
microfinance_data$Number_of_Institutions <- comma(microfinance_data$Number_of_Institutions)

kable(microfinance_data, col.names = c("Microfinance Institutions", "Number of Institutions"), caption = "Table 2. Microfinance Institutions and their numbers in the Philippines") %>%
  kable_styling(full_width = FALSE, position = "center") %>%
  add_footnote("*Source: Social Enterprise Development Partnerships, Inc., 2023 (Rapisura, 2023)*", notation = "none")
```

The role of MFIs in inclusion and poverty reduction plays an important role in the Philippines. [@cafamerica_microfinance_philippines_2024] further discussed the role of MFIs, such as providing financial services to low-income individuals and small businesses and promoting financial inclusion by ensuring marginalized communities have access to financial resources. The article explained further that providing financial resources and skills empowers individuals, especially women and vulnerable groups, strengthening social empowerment as a whole. Lastly, the authors also explained that microfinance helps in resilience building, particularly in disaster-prone regions, as MFIs assist communities in rebuilding by providing them with essential financial tools. 

There are various socioeconomic challenges that low-income households face in the Philippines that the MFIs must address. The lack of jobs and low pay are the challenges being faced. According to IBON Foundation executive director Sonny Africa, self-rated poverty continues to rise due to the lack of job opportunities combined with low wages and high inflation in the country [@africa2023filipinos].  Inflation is another challenge that low-income households face. [@ibon2024foodinflation] reported that the overall inflation and food inflation were at their highest during July 2024 highlighting the failure of government efforts to control inflation, which is contrary to the government's repeated claims of success. The article also mentioned that the PSA reported that the inflation for July 2024 increased to 4.4 percent from 3.7 percent in June 2024 after steadily increasing from 2.8 percent in January. The main drivers include higher electricity rates, more expensive meat and fruits, and an increase in oil prices. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)

inflation_data <- data.frame(
  Month = c("January", "June", "July"),
  Inflation_Rate = c(2.8, 3.7, 4.4)
)

kable(inflation_data, col.names = c("Month (2024)", "Inflation Rate (%)"), digits = 1, caption = "Table 3. Inflation Rate for the months of January, June, and July in 2024.") %>%
  kable_styling(full_width = FALSE, position = "center") %>%
  add_footnote("*Source: IBON Foundation, 2024*", notation = "none")

```

Additionally, the article discussed food inflation, which was at its worst last July 2024 for all groups. Food inflation was 3.3 percent in January 2024, then rose to 6.5 percent in June 2024, and is already at 6.7% by July 2024. Meanwhile, for the poorest 30% of  Filipino families who spend half of their incomes on food,  the food inflation rate was 5.3% in January 2024, rose to 8.4% in June 2024 then went up to 8.7% in July 2024.


```{r echo=FALSE, message=FALSE}
library(knitr)
library(kableExtra)
# Data frame with clear column names
food_inflation <- data.frame(
  Month_2024 = c("January", "June", "July"),
  Overall_Food_Inflation_Rate_Percent = c("2.8%", "3.7%", "4.4%"),
  Food_Inflation_Rate_Poorest_30_Percent = c("5.3%", "8.4%", "8.7%")
)

# Update column headers for display
colnames(food_inflation) <- c(
  "Month (2024)",
  "Overall Food Inflation Rate (%)",
  "Food Inflation Rate for the Poorest 30% (%)"
)

# Build the table
kable(food_inflation,
      caption = "Table 4: Food Inflation Rates in the Philippines for the Overall Population and the Poorest 30%, January, June, and July 2024",
      align = "lcc", booktabs = TRUE, format = "html") %>%
  kable_styling(full_width = FALSE, position = "center") %>%
  column_spec(2:3, width = "5cm") %>%
  add_footnote("*Source: IBON Foundation, 2024*", notation = "none")
```


Lastly, rising healthcare costs continue to negatively affect low-income households. The Global Medical Trends Survey of Wills Towers Watson (WTW) found that medical costs in the Philippines are projected to increase by double digits for the third consecutive year at 18.3 percent with little relief. The respondents of the WTW survey cited an increased use of health services, rising hospital and clinic costs, increasing professional fees, and higher frequency of diseases to be the main drivers of overall increased medical costs in the country [@wtw2025healthcare]. 

---
title: "Healthcare Benefit Cost Increases (2023–2025)"
output: html_document
---

```{r echo=FALSE, message=FALSE}
library(knitr)
library(kableExtra)

# Create the data frame
apac_costs <- data.frame(
  Country = c("Philippines"),
  `2023` = c("17.9%"),
  `2024` = c("19.3"),
  `2025` = c("18.3")
)

# Clean up column names to remove the "X"
colnames(apac_costs) <- c("Country", "2023", "2024", "2025 (Projected)")

# Build the table
kable(apac_costs,
      caption = "Table 5. Global medical trends: healthcare benefit cost increases, 2023 – 2025 (gross) - Philippines",
      align = "lccc", booktabs = TRUE, format = "html") %>%
  kable_styling(full_width = FALSE, position = "center") %>%
  column_spec(2:4, width = "3cm") %>%
  add_footnote("*Source: Global Medical Trends 2025 Survey of global health insurers - Wills   Towers Watsons, 2025*.", notation = "none")
```


## *2 Importance & Relevance*
The importance of studying MFIs in the Philippines is to have a better understanding of how MFIs improve the economic situation of low-income households through poverty alleviation and financial services from 2021 to 2024. The relevance of MFIs in the Philippines to the Sustainable Development Goals (SDGs) microfinance can help implement the SDGs.  [@srinivas2023channeling] explains that through providing undeserved populations financial access, Microfinance helps promote economic growth and reduce poverty, which are key components of several SDGs, such as SDG 1: No Poverty, SDG 8: Decent Work and Economic Growth, and SDG 9: Industry, Innovation and Infrastructure.  Additionally, the article discusses that microfinance supports other SDGs such as SDG 2: No Hunger, as it assists farmers and food producers in accessing credit,  and SDG 5: Gender Equality, as it gives help to women who may have otherwise been excluded from accessing formal financial services.


## *3 Objectives and Scope* 
#### *3.1 Objectives*
The objective of this study is to understand how effective microfinance institutions (MFIs) have been in assisting low-income households in improving their economic situation in the Philippines between 2021 and 2024. The study aims to measure the changes in household income among low-income MFI beneficiaries in the Philippines across four key periods: 2021 to 2024.  Furthermore, the study aims a to identify the most common purposes for which low-income households in the Philippines utilized microfinance loans and to analyze trends in loan repayment behavior among low-income MFI borrowers from 2021 to 2024. Finally, to compare the economic outcomes of low-income households with access to MFIs to those without access to MFIs during the period between 2021 and 2024. 

#### *3.2 Scope*
The scope of this study will focus on MFI beneficiaries from low-income households in the Philippines. The data of this study should span the years from 2021 to 2024. The study limits its focus to the utilization of microfinance loans and the repayment behavior of low-income households. Lastly, the study will focus on and discuss the available literature and data about MFIs in the Philippines. While this study aims to assess trends from 2021 to 2024, the existing literature is fragmented across individual years. As such, the literature review synthesizes findings from studies focusing on specific time frames within the period to provide a composite understanding of the broader trends. This limitation may affect the consistency of year-to-year comparisons.


## *4 Review of Related Literature* 

#### *4.1 Drivers of the market*

The Micro Finance market in the Philippines is developing as a result of the increased demand for financial inclusion, particularly among long-income households and small business owners. The growth of microfinance services is being supported by government programs and participation from global financial institutions. Access to microfinance products is being improved by the growing use of mobile banking and digital payment systems. Demand for microloans and financial services tailored to underprivileged groups is also being driven by the growth of small and medium-sized businesses (SMEs) and rural enterprises

```{r echo=FALSE, results='asis'}
library(knitr)
library(kableExtra)

# Create the table data
bank_data <- data.frame(
  Bank_Type = c("Rural Bank",
                "Universal Bank", "Thrift Bank", "Rural Bank", "Cooperative Bank"),
  Count = c(5, 1, 19, 96, 10),
  Group = c("Microfinance-Oriented Banks",
            rep("Microfinance-Engaged Banks", 4))
)

# Create the formatted table
bank_data |>
  kbl(
    caption = "Table _: List of Banks with Microfinance Operations",
    col.names = c("Bank Type", "Count", "Category"),
    align = "lcc"
  ) |>
  kable_styling(full_width = FALSE, position = "center") |>
  add_footnote("*Source: Bangko Sentral ng Pilipinas (BSP), 2024.*", notation = "none")
```

Microfinance institutions (MFIs) play a significant role in advancing financial inclusion in the Philippines, with a presence in 80% of municipalities and cities. According to the [@bsp_speech_2021], 9% of Filipino adults maintain accounts with MFIs, positioning them the third most preferred account provider after e-money issuers and banks. It was reported that 23% of Filipino borrowers sourced their loans from MFIs in 2021, surpassing government institutions, cooperatives, and lending-financing institutions. Despite the significant challenges posed by the COVID-19 pandemic, which caused a contraction in loan growth and a sharp decline in repayment rates, the microfinance sector has shown resilience, with loan growth rebounding and repayment rates recovering steadily by early 2021. 

MFIs, including 139 banks, microfinance NGOs with portfolios of ₱50.4 billion, and cooperatives with ₱288 billion in loans, enable microentrepreneurs to access credit without collateral and simple requirements, empowering many to “graduate” into formal banking. The BSP actively supports this sector by promoting credit access through guarantee programs and encouraging digitalization initiatives such as the Paleng-QR Ph and satellite technology for rural banks, which have proven essential during crises like typhoons 


#### *4.2 Changes in Household Income Among MFI beneficiaries*

The average annual income of Filipino families in 2023 was estimated at PhP 353.23 thousand. This was higher by 15.0 percent from the PhP 307.19 thousand average family income in 2021. Moreover, the average annual family income in 2018 was recorded at PhP 313.45 thousand.

On the other hand, Filipino families spent an average of PhP 258.05 thousand in 2023, an increase of 12.8 percent compared to the PhP 228.80 thousand expenditure in 2021. In 2018, the average annual family expenditure was registered at PhP 238.75 thousand. 


Link: https://psa.gov.ph/statistics/income-expenditure/fies

#### *4.3 Utilization of Microfinance Loans* 



#### *4.4 Poverty Reduction and Limitations of Microfinance in the Philippines*


## *Implications and Future Research* 

## *Conclusion*

## *References* 



