---
title: "**A Case Study on the Impact of Microfinance Institutions on Poverty Alleviation
  in Low-Income Households in the Philippines, 2021-2024**"
author: "U. Cortes, W.E. Limguangco"
date: "2025-06-10"
output: html_document
css: style.css
bibliography: references.bib
csl: apa.csl
---

## *Abstract* 

## *1 Background & Context* 
Poverty is one of the main issues the Philippines is experiencing to the present day. In 2021, the Philippine Statistics Authority (PSA)  (2022) reported that the poverty rate was 18.1 percent, equating to  19.99 million Filipinos, and the poverty incidence among Filipino families was 13.2 percent.  Furthermore, PSA reported that the proportion of Filipino families with insufficient income to meet basic food needs stood at 3.9 percent, and among Filipinos,  6 in 100 lived below the food threshold in 2021. In 2022, the PSA has not released the official poverty statistics of the Philippines for the year, however, according to Domingo (2022) of Inquirer, the poverty incidence in the Philippines was projected to ease to 17.1 percent from 18.1 percent in 2021, thanks to recovering labor market indicators according to the World Bank.

```{r echo=FALSE, results='asis'}
library(knitr)

# Data
years <- c("2021", "2022 (Estimation)", "2023", "2024 (Estimation)")
poverty_rates <- c(18.1, 17.1, 15.5, 13.6)

# Create table with custom column name
data <- data.frame(
  Year = years,
  check.names = FALSE  # Allows column names with spaces and symbols
)
data[["Poverty Incidence (%)"]] <- poverty_rates

# Display table
kable(data, caption = "Poverty Incidences in the Philippines from 2021–2024", align = "c")
```
The microfinance sector has experienced significant growth in the Philippines in recent years.  As of March 2022, the Bangko Sentral ng Pilipinas (BSP) reported that the microfinance sector has expanded to a loan portfolio of Php 406 billion (Batayola et al, 2024). There are approximately 6,183 microfinance institutions (MFIs) in the Philippines. The role of MFIs in inclusion and poverty reduction plays an important role in the Philippines. Quick and de Toledo (2024) further discussed the role of MFIs, such as providing financial services to low-income individuals and small businesses and promoting financial inclusion by ensuring marginalized communities have access to financial resources. Quick and Toledo explained further that providing financial resources and skills empowers individuals, especially women and vulnerable groups, strengthening social empowerment as a whole. Furthermore, the authors also explain that microfinance helps in resilience building, particularly in disaster-prone regions, as MFIs assist communities in rebuilding by providing them with essential financial tools. 


The microfinance sector has experienced significant growth in the Philippines in recent years.  As of March 2022, the Bangko Sentral ng Pilipinas (BSP) reported that the microfinance sector has expanded to a loan portfolio of Php 406 billion (Batayola et al, 2024). There are approximately 6,183 microfinance institutions (MFIs) in the Philippines. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
library(scales)

# Create the data
df <- data.frame(
  `Microfinance Institutions` = c(
    "Banks (as of 2022)",
    "Cooperatives (2020)",
    "Microfinance NGOs (as of 2022)",
    "Financing Companies (as of 2017)",
    "Lending Companies (as of 2017)",
    "Total"
  ),
  `Number of Institutions` = comma(c(139, 2762, 27, 673, 2582, 6183))
)

# Display the table
kable(df, align = "l", col.names = c("Microfinance Institutions", "Number of Institutions")) %>%
  kable_styling(full_width = FALSE, position = "center")
```

The role of MFIs in inclusion and poverty reduction plays an important role in the Philippines. Quick and de Toledo (2024) further discussed the role of MFIs, such as providing financial services to low-income individuals and small businesses and promoting financial inclusion by ensuring marginalized communities have access to financial resources. The article explained further that providing financial resources and skills empowers individuals, especially women and vulnerable groups, strengthening social empowerment as a whole. Lastly, the authors also explain that microfinance helps in resilience building, particularly in disaster-prone regions, as MFIs assist communities in rebuilding by providing them with essential financial tools. 

There are various socioeconomic challenges that low-income households face in the Philippines that the MFIs must address. The lack of jobs and low pay are the challenges being faced. According to IBON Foundation executive director Sonny Africa, self-rated poverty continues to rise due to the lack of job opportunities combined with low wages and high inflation in the country (ABS-CBN, 2023).  Inflation is another challenge that low-income households face. IBON Foundation (2024) reported that the overall inflation and food inflation were at their highest during July 2024 highlighting the failure of government efforts to control inflation, which is contrary to the government's repeated claims of success. IBON Foundation (2024) also mentioned that the PSA reported that the inflation for July 2024 increased to 4.4 percent from 3.7 percent in June 2024 after steadily increasing from 2.8 percent in January. The main drivers include higher electricity rates, more expensive meat and fruits, and an increase in oil prices. 

```{r} 

```



## *2 Importance & Relevance*
The importance of studying MFIs in the Philippines is to have a better understanding of how MFIs improve the economic situation of low-income households through poverty alleviation and financial services from 2021 to 2024. The relevance of MFIs in the Philippines to the Sustainable Development Goals (SDGs) microfinance can help implement the SDGs.  Srinivas (2024) explains that through providing undeserved populations financial access, Microfinance helps promote economic growth and reduce poverty, which are key components of several SDGs, such as SDG 1: No Poverty, SDG 8: Decent Work and Economic Growth, and SDG 9: Industry, Innovation and Infrastructure.  Furthermore, Srivinivas (2024) states that microfinance supports other SDGs such as SDG 2: No Hunger, as it assists farmers and food producers in accessing credit,  and SDG 5: Gender Equality, for it gives help to women who may have otherwise been excluded from accessing formal financial services.


## *3 Objectives and Scope* 
#### *3.1 Objectives*
The objective of this study is to understand how effective microfinance institutions (MFIs) have been in assisting low-income households in improving their economic situation in the Philippines between 2021 and 2024. The study aims to measure the changes in household income among low-income MFI beneficiaries in the Philippines across four key periods: 2021 to 2024.  Furthermore, the study aims a to identify the most common purposes for which low-income households in the Philippines utilized microfinance loans and to analyze trends in loan repayment behavior among low-income MFI borrowers from 2021 to 2024. Finally, to compare the economic outcomes of low-income households with access to MFIs to those without access to MFIs during the period between 2021 and 2024. 

#### *3.2 Scope*
The scope of this study will focus on MFI beneficiaries from low-income households in the Philippines. The data of this study should span the years from 2021 to 2024. The study limits its focus to the utilization of microfinance loans and the repayment behavior of low-income households. Lastly, the study will focus on the economic outcomes of having access to MFIs in the Philippines.



## *4 Review of Related Literature* 

#### *4.1 Drivers of the market*

The Micro Finance market in the Philippines is developing as a result of the increased demand for financial inclusion, particularly among long-income households and small business owners. The growth of microfinance services is being supported by government programs and participation from global financial institutions. Access to microfinance products is being improved by the growing use of mobile banking and digital payment systems. Demand for microloans and financial services tailored to underprivileged groups is also being driven by the growth of small and medium-sized businesses (SMEs) and rural enterprises

```{r echo=FALSE, results='asis'}
library(knitr)
library(kableExtra)

# Create the table data
bank_data <- data.frame(
  Bank_Type = c("Rural Bank",
                "Universal Bank", "Thrift Bank", "Rural Bank", "Cooperative Bank"),
  Count = c(5, 1, 19, 96, 10),
  Group = c("Microfinance-Oriented Banks",
            rep("Microfinance-Engaged Banks", 4))
)

# Create the formatted table
bank_data |>
  kbl(
    caption = "Table _: List of Banks with Microfinance Operations",
    col.names = c("Bank Type", "Count", "Category"),
    align = "lcc"
  ) |>
  kable_styling(full_width = FALSE, position = "center") |>
  add_footnote("*Source: Bangko Sentral ng Pilipinas (BSP), 2024.*", notation = "none")
```

Microfinance institutions (MFIs) play a significant role in advancing financial inclusion in the Philippines, with a presence in 80% of municipalities and cities. According to the [@bsp_speech_2021], 9% of Filipino adults maintain accounts with MFIs, positioning them the third most preferred account provider after e-money issuers and banks. It was reported that 23% of Filipino borrowers sourced their loans from MFIs in 2021, surpassing government institutions, cooperatives, and lending-financing institutions. Despite the significant challenges posed by the COVID-19 pandemic, which caused a contraction in loan growth and a sharp decline in repayment rates, the microfinance sector has shown resilience, with loan growth rebounding and repayment rates recovering steadily by early 2021. 

MFIs, including 139 banks, microfinance NGOs with portfolios of ₱50.4 billion, and cooperatives with ₱288 billion in loans, enable microentrepreneurs to access credit without collateral and simple requirements, empowering many to “graduate” into formal banking. The BSP actively supports this sector by promoting credit access through guarantee programs and encouraging digitalization initiatives such as the Paleng-QR Ph and satellite technology for rural banks, which have proven essential during crises like typhoons 


#### *4.2 Changes in Household Income Among MFI beneficiaries*

The average annual income of Filipino families in 2023 was estimated at PhP 353.23 thousand. This was higher by 15.0 percent from the PhP 307.19 thousand average family income in 2021. Moreover, the average annual family income in 2018 was recorded at PhP 313.45 thousand.

On the other hand, Filipino families spent an average of PhP 258.05 thousand in 2023, an increase of 12.8 percent compared to the PhP 228.80 thousand expenditure in 2021. In 2018, the average annual family expenditure was registered at PhP 238.75 thousand. 


Link: https://psa.gov.ph/statistics/income-expenditure/fies

#### *4.3 Utilization of Microfinance Loans* 



#### *4.4 Poverty Reduction and Limitations of Microfinance in the Philippines*


## *Implications and Future Research* 

## *Conclusion*

## *References* 



